---
output:
  knitrBootstrap::bootstrap_document:
    theme.chooser: FALSE
    highlight.chooser: FALSE
  html_document:
    toc: true
    theme: spacelab
---

Memory issues with BiocParallel
=====

This repo is about a memory issue with `BiocParallel` in the computing cluster I have access to managed by a SGE system. The full post can be found at the [Bioc-devel](https://stat.ethz.ch/pipermail/bioc-devel/2015-July/007788.html) mailing list. Basically, it seems that after R changed to version 3.2 -- which matches other major changes to `BiocParallel` -- the memory used greatly increased when using `SnowParam()`. 

I first noticed this issue when attempting to reproduce the results of [step1-fullCoverage.sh](https://github.com/leekgroup/derSoftware/blob/gh-pages/step1-fullCoverage.sh) using R 3.2.x (using `sh step1-fullCoverage.sh brainspan`). Previously with R 3.1.x, the script used 173.5 GB of RAM and in R 3.2 and 3.2.x it reached 450 GB before getting shut down by the cluster system. The main package used in that script is basically identical between R versions 3.1.x and 3.2, and while `SnowParam()` did change it's also likely that the issue is related to another package(s). Note that the results and memory used are reproducible with R 3.1.x.

Here I made some scripts to try to show the problem with a smaller data set. The actual scripts are [SnowParam-memory.R](https://github.com/lcolladotor/SnowParam-memory/tree/gh-pages/SnowParam-memory.R) and [SnowParam-memory.sh](https://github.com/lcolladotor/SnowParam-memory/tree/gh-pages/SnowParam-memory.sh) which are run using `sh SnowParam-memory.sh`. It runs the same code for R versions 3.1.x, 3.2 and 3.2.x. 


## Data

Below is the summary information for these jobs with memory shown in GB used, as reported by the cluster I'm using. The actual files are at [mem_emails/](https://github.com/lcolladotor/SnowParam-memory/tree/gh-pages/mem_emails). The log files are available at [logs/](https://github.com/lcolladotor/SnowParam-memory/tree/gh-pages/logs).

For example, [mem_emails/der-snow-3.1.x.txt](https://github.com/lcolladotor/SnowParam-memory/blob/gh-pages/mem_emails/der-snow-3.1.x.txt) has an entry with the email information for the job per replicate run. The email information includes among other things, the maximum memory as measured by the cluster under "Max vmem". Each entry starts with the job name, for example [job 6416599](https://github.com/lcolladotor/SnowParam-memory/blob/gh-pages/mem_emails/der-snow-3.1.x.txt#L1). The job number is then useful to identify which are the corresponding log files. The log files will have names with extentions __.e__ and __.o__ for stderr and stdout respectively, and with the same digits as the job id. So for this example, the log files are [logs/der-snow-3.1.x.e6416599](https://github.com/lcolladotor/SnowParam-memory/blob/gh-pages/logs/der-snow-3.1.x.e6416599) and [logs/der-snow-3.1.x.o6416599](https://github.com/lcolladotor/SnowParam-memory/blob/gh-pages/logs/der-snow-3.1.x.o6416599).

```{r 'summary', bootstrap.show.code = FALSE}
df <- data.frame(
    memory = c(
        6.404, 5.170, 5.036,
        12.926, 12.899, 11.735,
        1.573, 1.573, 1.508,
        7.286, 7.285, 5.036,
        12.370, 12.303, 11.676,
        1.574, 1.574, 1.508,
        7.286, 7.285, 5.036,
        12.394, 12.452, 11.974,
        1.574, 1.574, 1.508,
        7.286, 7.285, 5.036,
        12.161, 12.422, 11.974,
        1.306, 1.306, 1.210
    ),
    R = factor(rep(c('3.2.x', '3.2', '3.1.x'), 3),
        levels = c('3.1.x', '3.2', '3.2.x')),
    param = rep(c('snow', 'multicore', 'serial'), each = 3),
    run = factor(rep(1:4, each = 9))
)
knitr::kable(df, format = 'html')
```

The same information is shown below in two plots. The first one use _loess_ to summarize the overall trend while the second one shows one line per each replicate run.

```{r 'plot', bootstrap.show.code = FALSE, warning = FALSE}
library('ggplot2')
ggplot(df, aes(y = memory, x = R, colour = param, shape = run, group = param)) + geom_point() + geom_smooth(method = loess, se = FALSE)

## One line per run
ggplot(df, aes(y = memory, x = R, colour = param, shape = run, group = paste0(df$param, df$run))) + geom_point() + geom_line()
```

Note that starting in R 3.2, the type of cluster is SOCK instead of PSOCK. I ignore if this could explain the observed difference.

## `derfinder` example

[SnowParam-memory-derfinder.R](https://github.com/lcolladotor/SnowParam-memory/tree/gh-pages/SnowParam-memory.R) and [SnowParam-memory-derfinder.sh](https://github.com/lcolladotor/SnowParam-memory/tree/gh-pages/SnowParam-memory.sh) compose a second example using objects and some of the code from the original use case. I wrote to see if it lead to the same memory fold changes I observe with my analysis script. Here are some of the results.

```{r 'summary-der', bootstrap.show.code = FALSE}
df2 <- data.frame(
    memory = c(
        13.091, 12.308, 7.175,
        13.856, 13.268, 8.475,
        1.027, 1.040, 927.723 / 1024,
        12.871, 10.904, 7.174,
        13.671, 13.789, 8.475,
        1.029, 1.042, 927.723 / 1024,
        12.871, 10.940, 7.174,
        13.675, 13.793, 8.465,
        1.028, 1.042, 931.301 / 1024,
        12.866, 11.072, 7.175,
        13.671, 13.793, 8.473,
        1.011, 1.030, 929.055 / 1024
    ),
    R = factor(rep(c('3.2.x', '3.2', '3.1.x'), 3),
        levels = c('3.1.x', '3.2', '3.2.x')),
    param = rep(c('snow', 'multicore', 'serial'), each = 3),
    run = factor(rep(1:4, each = 9))
)
knitr::kable(df2, format = 'html', digits = 3)
```

The same information is shown below in two plots. The first one use _loess_ to summarize the overall trend while the second one shows one line per each replicate run.

```{r 'plot-der', bootstrap.show.code = FALSE, warning = FALSE}
ggplot(df2, aes(y = memory, x = R, colour = param, shape = run, group = param)) + geom_point() + geom_smooth(method = loess, se = FALSE)

## One line per run
ggplot(df2, aes(y = memory, x = R, colour = param, shape = run, group = paste0(df2$param, df2$run))) + geom_point() + geom_line()
```

In this example, the memory mean fold change when using `SnowParam()` between R 3.2.x and 3.1.x is `r signif(mean(df2$memory[c(1, 10, 19, 28)]) / mean(df2$memory[c(3, 12, 21, 30)]), 3)`x, compared to `r signif(mean(df$memory[c(1, 10, 19, 28)]) / mean(df$memory[c(3, 12, 21, 30)]), 3)`x from the first example. In comparison, with `SerialParam()` the same mean fold changes are `r signif(mean(df2$memory[c(7, 16, 25, 34)]) / mean(df2$memory[c(9, 18, 27, 36)]), 3)`x and `r signif(mean(df$memory[c(7, 16, 25, 34)]) / mean(df$memory[c(9, 18, 27, 36)]), 3)`x respectively.

As shown in the [compare view](https://github.com/lcolladotor/derfinder/compare/bioc3.0...master) nothing changed in [filterData()](https://github.com/lcolladotor/derfinder/blob/master/R/filterData.R) between R 3.1.x and 3.2.x.


This page was last updated at `r Sys.time()`.

